Manuscript

### 2022-03-28 analysis/manuscript.Rmd
wflow_publish("analysis/manuscript.Rmd")

tar_load(couple_MR_vs_trait_corr)
tar_load(corr_impact_by_PCs)
tar_load(corr_impact_by_traits)
tar_load(corr_impact_by_coords)
tar_load(corr_potential_trait_confounders)
tar_load(traits_corr)
tar_load(PCs_corr)
tar_load(traits_corr2_filled)
tar_load(household_MR_summary_AM)
tar_load(num_tests_by_PCs)
tar_load(num_tests_by_PCs_AM_sig)
tar_load(household_MR_binned_het_sex_spec)
tar_load(household_MR_binned_het_SNPmeta)
tar_load(traits_final)
tar_load(household_MR_AM_FvsM_fig)
tar_load(confounder_traits)
tar_load(proxyMR_prod_comparison_fig_yiyp_adj)
tar_load(proxyMR_comparison_summary_yiyp_adj_SNPmeta)
tar_load(household_MR_summary_AM)
tar_load(household_MR_binned_het_SNPmeta)
tar_load(potential_trait_confounders_MVMR)
tar_load(corr_mat_traits)

@@

```{r pearson_vs_spearman, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE, warning = FALSE}
tar_load(Neale_SGG_dir_filt)
tar_load(traits_corr_spearman)
tar_load(traits_corr)

@@

pearson_vs_spearman_dat <- inner_join(traits_corr, unique(Neale_SGG_dir_filt_unique[,c("sgg_phesant_name", "variable_type")]), by = c("ID"="sgg_phesant_name")) %>% dplyr::rename(r2_pearson = r2) %>%
  inner_join(traits_corr_spearman[,c("ID", "r2")]) %>% dplyr::rename(r2_spearman = r2) %>% 
  as_tibble() %>% type_convert() %>%
  mutate(r_pearson = sqrt(r2_pearson)) %>% 
  mutate(r_spearman = sqrt(r2_spearman))

pearson_vs_spearman_plot <- ggplot2::ggplot(data = pearson_vs_spearman_dat, 
                                            ggplot2::aes(x = r_pearson,
                                                         y = r_spearman, 
                                                         color = factor(variable_type))) + 
  geom_point() +
  theme_minimal() +
  ggplot2::labs(x = paste0("Pearson correlation"),
                y = paste("Spearman correlation")) + 
  scale_color_manual(name = "Variable type", values = custom_col[c(1,2,3)], labels=c("Binary","Continuous (IRNT)", "Ordinal"))

@@

Sought to compare the raw correlation amongst couples vs. the standardized MR effects. There are many traits where the `correlation > MR effects` including some traits of interest such as standing height, place of birth, among many others.

@@

The table below displays the `r dim(corr_vs_MR_effects_sig_dat.DT)[1]` traits that have correlation significantly different from the MR-estimate. Of the significant difference `r table(corr_vs_MR_effects_sig_dat$correlation_larger)[[1]]` have MR-estimates larger than than correlation, suggesting the presence of a negative confounder.

@@

1. Restrict to only traits that have correlation significantly different from MR-estimate (`r dim(corr_vs_MR_effects_sig_dat)[1]` traits have correlation different from MR_estimate).
2. For each of these traits, find the traits which have a causal effect on this trait (MR p-val < 0.05/[number of effective tests]).
3. Calculate the correlation due to confounding as: $C = (\alpha_{z\rightarrow x})^2 * \alpha_{z_i\rightarrow z_p}$.
4. Calculate the ratio of this correlation ($C$) and the correlation of $X$ in partners ($C/r_x$). 
5. Remove any `potential_counders` with correlation with `trait_ID` > `r corr_trait_threshold`.
6. For each `trait_ID` prune remaining `potential_counders` so that all have correlations <  `r corr_trait_threshold`, prioritized by their `corr_due_to_confounding_ratio`.

A summary of the top 5 confounders for each `trait_ID` can be seen below:

```{r simple_C_model, echo = FALSE}
library(knitr)
knitr::include_graphics("assets/simple_confounder_model.png", error = FALSE)

@@

number_of_potential_confounders <- dim(corr_potential_trait_confounders)[1]/dim(corr_vs_MR_effects_sig_dat)[1]

non_sig_confounders <- corr_potential_trait_confounders %>% filter(!sig_confounder) %>% group_by(outcome_ID) %>%
  dplyr::summarize(description = outcome_description[1], n = n()) %>% 
  filter(n==number_of_potential_confounders) %>%
  mutate(n = n-number_of_potential_confounders) %>% mutate(potential_confounders = NA) %>%
  mutate(mean_C_ratio = NA) %>%
  dplyr::rename(trait_ID = outcome_ID)

sig_confounders <- corr_potential_trait_confounders %>% filter(sig_confounder) %>% group_by(outcome_ID) %>%
  arrange(-corr_due_to_confounding_ratio) %>%
  mutate(total_potential_confounders = n()) %>%
  dplyr::summarize(description = outcome_description[1],
                   total_potential_confounders = total_potential_confounders[1],
                   potential_confounders = paste0(exposure_description, collapse = "; "), 
                   max_C = max(corr_due_to_confounding), 
                   min_C = min(corr_due_to_confounding)) %>%
  dplyr::rename(trait_ID = outcome_ID) %>%
  dplyr::rename(trait_description = description) %>% 
  dplyr::select(trait_description, trait_ID, everything())

sig_confounders_top5 <- corr_potential_trait_confounders %>% filter(sig_confounder) %>% group_by(outcome_ID) %>%
  arrange(-corr_due_to_confounding_ratio) %>%
  mutate(total_potential_confounders = n()) %>%
  slice_head(n = 5) %>%
  #mutate(potential_confounders = paste0(exposure_description, collapse = "; ")) %>%
  dplyr::summarize(trait_description = outcome_description[1],
                   total_potential_confounders = total_potential_confounders[1],
                   top5_potential_confounders = paste0(exposure_description, collapse = "; "), 
                   max_C = max(corr_due_to_confounding), 
                   min_C = min(corr_due_to_confounding)) %>%
  dplyr::rename(trait_ID = outcome_ID) %>%
  left_join(corr_vs_MR_effects_sig_dat.DT, by = c("trait_description", "trait_ID")) %>%
  dplyr::select(trait_description, trait_ID, couple_r, IVW_beta, IVW_pval,everything())

@@

max_C_data <- corr_potential_trait_confounders %>% filter(sig_confounder) %>% group_by(outcome_ID) %>%
  arrange(-corr_due_to_confounding_ratio) %>%
  mutate(total_potential_confounders = n()) %>%
  slice_head(n = 1) %>%
  left_join(couple_MR_vs_trait_corr %>% 
              dplyr::rename(outcome_ID_IVW_beta = IVW_beta) %>%
              dplyr::rename(outcome_ID_IVW_se = IVW_se) %>%
              dplyr::rename(outcome_ID_couple_r = couple_r) %>% 
              dplyr::rename(outcome_ID_couple_r_se = couple_r_se) %>% 
              dplyr::select(outcome_ID, outcome_ID_IVW_beta, outcome_ID_IVW_se, outcome_ID_couple_r, outcome_ID_couple_r_se)) %>%
  mutate(same_trait_corr_minus_MR_est = abs(outcome_ID_couple_r)- abs(outcome_ID_IVW_beta)) %>%

  dplyr::rename(Z_beta = IVW_beta) %>%
  dplyr::rename(Z_se = IVW_se) %>%
  dplyr::rename(Z_pval = IVW_pval) %>%
  mutate(Z_sq_beta = Z_beta^2) %>%
  mutate(Z_sq_se = sqrt(4*Z_beta^2*Z_se^2+2*Z_se^4)) %>%
  dplyr::rename(Z_AM_beta = exposure_ID_AM_IVW_beta) %>%
  dplyr::rename(Z_AM_se = exposure_ID_AM_IVW_se) %>%
  mutate(Z_prod = Z_sq_beta*Z_AM_beta) %>%
  mutate(Z_prod_var = variance_of_product(Z_sq_beta, Z_sq_se, Z_AM_beta, Z_AM_se)) %>%
  mutate(Z_prod_se = sqrt(Z_prod_var)) %>%
  mutate(same_trait_corr_minus_MR_est_se = variance_of_sum(outcome_ID_IVW_se, outcome_ID_couple_r_se))

@@

  theme_minimal() +
  ggplot2::labs(x = paste0("max C"),
                y = paste("abs(correlation) - abs(MR-estimate)")) +
  theme(legend.title=element_blank()) +
  #geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  theme_ipsum()

@@

Finally, we computed a MVMR to sum up the effect of all potential confounders as shown in the DAG below and compared these C-statistics to the difference between phenotypic correlation and MR-estimates.

@@

```{r MVMR_C_plot, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE}

discrepentant_traits <- unique(corr_potential_trait_confounders$outcome_ID)

@@

  corr_potential_trait_confounders_i <- corr_potential_trait_confounders %>% filter(outcome_ID == !!outcome_ID)

@@

  MVMR_result_org <- MVMR_result_org %>% mutate(beta_2_var = 4*beta^2*se^2+2*se^4) %>%
    mutate(beta_2_se = sqrt(beta_2_var)) %>% 
    left_join(corr_potential_trait_confounders_i %>% dplyr::select(exposure_ID, exposure_ID_AM_IVW_beta, exposure_ID_AM_IVW_se), by = c("exposure" = "exposure_ID")) %>%
    dplyr::rename(MVMR_Z_beta = beta) %>%
    dplyr::rename(MVMR_Z_se = se) %>%
    dplyr::rename(MVMR_Z_pval = pval) %>%
    mutate(MVMR_Z_sq_beta = MVMR_Z_beta^2) %>%
    dplyr::rename(MVMR_Z_sq_se = beta_2_se) %>% 
    dplyr::rename(Z_AM_beta = exposure_ID_AM_IVW_beta) %>%
    dplyr::rename(Z_AM_se = exposure_ID_AM_IVW_se) %>%
    mutate(Z_prod = MVMR_Z_sq_beta*Z_AM_beta) %>%
    mutate(Z_prod_var = variance_of_product(MVMR_Z_sq_beta, MVMR_Z_sq_se, Z_AM_beta, Z_AM_se))

@@

@@

AM_result_sig <- household_MR_summary_AM %>% filter(IVW_pval < 0.05/num_tests_by_PCs) 
AM_sig_traits <-household_MR_summary_AM %>% filter(IVW_pval < 0.05/num_tests_by_PCs) %>% pull(exposure_ID) %>% unique()
AM_sex_spec_result_sig <- household_MR_binned_het_sex_spec %>% filter(same_trait ==TRUE) %>% filter(exposure_ID %in% AM_sig_traits) %>% filter(sex_het_pval < 0.05/num_tests_by_PCs_AM_sig)

@@

- For each trait in the pipeline, we tested the causal effect within couples (i.e. the number of analyses corresponds to the number of traits (`r dim(traits_final)[1]`)). 
- The significant results after adjusting for multiple hypothesis testing are shown below. 
- `IVW_meta_beta` and `IVW_meta_pval` corresponds to the beta and p-value of the meta-analyzed MR across sexes, respectively (i.e. MR estimates were computed in each sex-seperately using sex-specific SNP-exposure and SNP-outcome results and then meta-analyzed).  
- After adjusting for multiple tests (p < `0.05/`r num_tests_by_PCs``), identified `r dim(AM_result_sig)[1]` significant assortative mating (same-trait) MR results (corresponding to the table below)..

```{r, AM_sig, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), echo = FALSE, include = TRUE}

@@

AM_result_sig_nom.DT <- AM_sig_sex_het_nom %>% dplyr::select(trait_description, trait_ID, exposure_sex, IVW_beta, IVW_se, IVW_pval, sex_het_pval) %>% unique()
```

@@

- For each AM MR ($X_{p} \sim X_i$), analyses were run in each sex separately. 
- Of the `r dim(AM_sig_traits)[1]` significant results shown above, we tested to see if there was any difference in AM MR estimates between sexes. 
- A similar adjustment for multiple hypothesis testing was applied as above, resulting in `r num_tests_by_PCs_AM_sig` number of effective tests. 
- After adjusting for multiple hypothesis testing (p < `0.05/`r num_tests_by_PCs_AM_sig``), `r dim(AM_sex_spec_result_sig)[1]` traits showed significant differences amongst sexes.
- The table below shows the *nominally significant* results at `p < 0.05` (of which there are `r dim(AM_result_sig_nom.DT)[1]/2`).

@@

nominally_sig_sex_differences <- AM_result_sig_nom.DT %>% pivot_wider(names_from = exposure_sex, values_from = c(IVW_beta, IVW_se, IVW_pval)) 
# write.table(nominally_sig_sex_differences, "docs/manuscript_tables/nominally_sig_sex_differences.txt", row.names = F, sep=",", quote = F)
```

@@

- For each AM MR ($X_{p} \sim X_i$), analyses were run in the full sample as well as in 5 roughly equal sized bins according to time couple had been together (`time_together_even_bins`) estimated using [time at household variable](https://biobank.ctsu.ox.ac.uk/showcase/field.cgi?id=699), and median age (`age_even_bins`). 
- Of the `r length(AM_sig_traits)` significant results shown above, we tested to see if there was any significant difference in AM MR estimates amongst the two grouping variables. 
- Specifically, the outcome data set was split into 5 bins, and the SNP-outcome effect was estimated in each bin separately (and each sex seperately). 
- These outcome-SNP effects were then used to generate bin-specific MR estimates, using the same SNP-exposure effects from Neale. 
- To estimate the difference among bins, we used two approaches: 
    1. The Cochran's Q test to test for heterogeneity in meta-analyses *(of which we obtained no significant results, so not presenting in paper)*.
    2. Tested the significance of the slope of linear model of median bin (either `age` or `time together`) versus the bin-specific MR estimate.

@@

We also tested the phenotypic correlation among the different bins. Below are the traits that show a significant trend either with Pearson's or Spearman's correlation (p < `r 0.05/num_tests_by_PCs`).

@@

tar_load(binned_pheno_corrs)

@@

binned_pheno_corr_sig <- binned_pheno_corr_summary %>% filter(bin_slope_pval_spearman < 0.05/num_tests_by_PCs | bin_slope_pval_pearson < 0.05/num_tests_by_PCs)

binned_pheno_corr_sig.DT <- replace_Neale_ID(binned_pheno_corr_sig, traits_corr2_filled, "trait", "trait_description") %>%
  dplyr::rename(trait_ID = trait) %>%
  dplyr::select(trait_description, trait_ID, everything())

format_round_cols <- colnames(dplyr::select_if(binned_pheno_corr_sig.DT, is.numeric))

DT::datatable(binned_pheno_corr_sig.DT, width = "100%") %>% formatRound(columns=format_round_cols, digits = 3)

# write.table(binned_pheno_corr_sig.DT, "docs/manuscript_tables/binned_phenotypic_correlations.txt", row.names = F, sep=",", quote = F)

@@

# Heterogeneity stats.

@@

We tested the impact of confounding on genetic PCs, N/E coordinates and the following 4 traits:

@@

```{r corr_due_to_confounding_ratio, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), echo = FALSE}

@@

>>>>

tar_load(Neale_SGG_dir_filt)
tar_load(traits_corr_spearman)
tar_load(couple_MR_vs_trait_corr)
tar_load(corr_impact_by_PCs)
tar_load(corr_impact_by_traits)
tar_load(corr_impact_by_coords)
tar_load(corr_potential_trait_confounders_pos)
tar_load(traits_corr)
tar_load(PCs_corr)
tar_load(traits_corr2_filled)
tar_load(household_MR_summary_AM)
tar_load(num_tests_by_PCs)
tar_load(num_tests_by_PCs_AM_sig)
tar_load(household_MR_binned_het_sex_spec)
tar_load(household_MR_binned_het_SNPmeta)
tar_load(traits_final)
tar_load(household_MR_AM_FvsM_fig)
tar_load(confounder_traits)
tar_load(proxyMR_prod_comparison_fig_yiyp_adj)
tar_load(proxyMR_comparison_summary_yiyp_adj_SNPmeta)
tar_load(household_MR_summary_AM)
tar_load(household_MR_binned_het_SNPmeta)
#tar_load(potential_trait_confounders_MVMR)
tar_load(corr_mat_traits)
tar_load(corr_impact_by_traits_causal)
tar_load(household_MR_SNPmeta_sensitivity)
tar_load(binned_pheno_corrs)

@@

```{r pearson_vs_spearman, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE, warning = FALSE}

@@

## looks at all traits *before* the correlation filter

pearson_vs_spearman_dat <-
  inner_join(traits_corr,
             unique(Neale_SGG_dir_filt_unique[, c("sgg_phesant_name", "variable_type")]),
             by = c("ID" = "sgg_phesant_name")) %>%
  dplyr::rename(r2_pearson = r2) %>%
  inner_join(traits_corr_spearman[, c("ID", "r2")]) %>% dplyr::rename(r2_spearman = r2) %>%
  as_tibble() %>% type_convert() %>%
  mutate(r_pearson = sqrt(r2_pearson)) %>%
  mutate(r_spearman = sqrt(r2_spearman))

pearson_vs_spearman_plot <-
  ggplot2::ggplot(data = pearson_vs_spearman_dat,
                  ggplot2::aes(
                    x = r_pearson,
                    y = r_spearman,
                    color = factor(variable_type)
                  )) +
  geom_point() +
  theme_minimal() +
  ggplot2::labs(x = paste0("Pearson correlation"),
                y = paste("Spearman correlation")) +
  scale_color_manual(
    name = "Variable type",
    values = custom_col[c(1, 2, 3)],
    labels = c("Binary", "Continuous (IRNT)", "Ordinal")
  )

@@

We sought to compare the raw correlation amongst couples vs. the standardized MR effects. There are many traits where the `correlation > MR effects` including some traits of interest such as standing height, place of birth, among many others.

@@

The table below displays the `r dim(corr_vs_MR_effects_sig_dat.DT)[1]` traits that have correlation significantly different from the MR-estimate. Of the significant difference `r table(corr_vs_MR_effects_sig_dat$correlation_larger)[[1]]` have MR-estimates larger than than correlation, suggesting the presence of a negative confounder (this table corresponds to "sig_corr_vs_MR" in the excel document).

@@

Sought to identify potential confounders (from our list of traits in the pipeline) which may, in part, explain the discrepant estimates, as follows:

1. Restrict to only traits that have correlation significantly different from MR-estimate (`r dim(corr_vs_MR_effects_sig_dat)[1]` traits have correlation different from MR_estimate).
1. Remove any `potential_counders` with correlation with `trait_ID` > `r corr_trait_threshold`.
1. For each of the remaining $Z$ `potential_counders` for each trait $X$, find the potential confounders which have a causal effect on trait $X$ ($\alpha_{z\rightarrow x}$ p-val < 0.05/[number of effective tests]).
1. Split results into those cases where MR-estimate is **greater** than the phenotypic correlation and those that are **less** than the correlation. 
1. For traits where correlation was less than MR-estimate, we restricted to negative confounders (i.e. negative $\alpha_{z_i\rightarrow z_p}$), and conversely for traits where the correlation was greater than the MR-estimate we restricted to positive confounders (i.e. positive $\alpha_{z_i\rightarrow z_p}$) [note that only one significant MR-effect is negative].  
1. Of the remaining $Z$, we further filtered to those with a significant within couple MR effect ($\alpha_{z_i\rightarrow z_p}$ p < 0.05/[number of remaining $Z$]). 
1. Calculate the correlation due to confounding as: $C = (\alpha_{z\rightarrow x})^2 * \alpha_{z_i\rightarrow z_p}$.
1. Calculate the ratio of this correlation ($C$) and the correlation of $X$ in partners ($C/r_x$). 
1. For each `trait_ID` prune remaining `potential_counders` so that all have correlations <  `r corr_trait_threshold`, prioritized by their `corr_due_to_confounding_ratio`.

```{r simple_C_model, echo = FALSE}
library(knitr)
knitr::include_graphics("assets/simple_confounder_model.png", error = FALSE)
```

A summary of the top 5 confounders for each `trait_ID` can be seen below (a variation of this table, include *all* potential confounders is included in the excel sheet titled "sig_confounders"):

@@

number_of_potential_confounders <- dim(corr_potential_trait_confounders_pos)[1]/dim(corr_vs_MR_effects_sig_dat)[1]

non_sig_confounders <- corr_potential_trait_confounders_pos %>% filter(!sig_confounder) %>% group_by(outcome_ID) %>%
  dplyr::summarize(description = outcome_description[1], n = n()) %>% 
  filter(n==number_of_potential_confounders) %>%
  mutate(n = n-number_of_potential_confounders) %>% mutate(potential_confounders = NA) %>%
  mutate(mean_C_ratio = NA) %>%
  dplyr::rename(trait_ID = outcome_ID)

sig_confounders <- corr_potential_trait_confounders_pos %>% filter(sig_confounder) %>% group_by(outcome_ID) %>%
  arrange(-corr_due_to_confounding_ratio) %>%
  mutate(total_potential_confounders = n()) %>%
  dplyr::summarize(description = outcome_description[1],
                   total_potential_confounders = total_potential_confounders[1],
                   potential_confounders = paste0(exposure_description, collapse = "; "), 
                   max_C = max(corr_due_to_confounding), 
                   min_C = min(corr_due_to_confounding)) %>%
  dplyr::rename(trait_ID = outcome_ID) %>%
  dplyr::rename(trait_description = description) %>% 
  dplyr::select(trait_description, trait_ID, everything())

sig_confounders_top5 <- corr_potential_trait_confounders_pos %>% filter(sig_confounder) %>% group_by(outcome_ID) %>%
  arrange(-corr_due_to_confounding_ratio) %>%
  mutate(total_potential_confounders = n()) %>%
  slice_head(n = 5) %>%
  #mutate(potential_confounders = paste0(exposure_description, collapse = "; ")) %>%
  dplyr::summarize(trait_description = outcome_description[1],
                   total_potential_confounders = total_potential_confounders[1],
                   top5_potential_confounders = paste0(exposure_description, collapse = "; "), 
                   max_C = max(corr_due_to_confounding), 
                   min_C = min(corr_due_to_confounding)) %>%
  dplyr::rename(trait_ID = outcome_ID) %>%
  left_join(corr_vs_MR_effects_sig_dat.DT, by = c("trait_description", "trait_ID")) %>%
  dplyr::select(trait_description, trait_ID, couple_r, IVW_beta, IVW_pval,everything())

@@

max_C_data <- corr_potential_trait_confounders_pos %>% filter(sig_confounder) %>% group_by(outcome_ID) %>%
  arrange(-corr_due_to_confounding_ratio) %>%
  mutate(total_potential_confounders = n()) %>%
  slice_head(n = 1) %>%
  left_join(couple_MR_vs_trait_corr %>% 
              dplyr::rename(outcome_ID_IVW_beta = IVW_beta) %>%
              dplyr::rename(outcome_ID_IVW_se = IVW_se) %>%
              dplyr::rename(outcome_ID_couple_r = couple_r) %>% 
              dplyr::rename(outcome_ID_couple_r_se = couple_r_se) %>% 
              dplyr::select(outcome_ID, outcome_ID_IVW_beta, outcome_ID_IVW_se, outcome_ID_couple_r, outcome_ID_couple_r_se)) %>%
  mutate(same_trait_corr_minus_MR_est = abs(outcome_ID_couple_r)- abs(outcome_ID_IVW_beta)) %>%

  dplyr::rename(Z_beta = IVW_beta) %>%
  dplyr::rename(Z_se = IVW_se) %>%
  dplyr::rename(Z_pval = IVW_pval) %>%
  mutate(Z_sq_beta = Z_beta^2) %>%
  mutate(Z_sq_se = sqrt(4*Z_beta^2*Z_se^2+2*Z_se^4)) %>%
  dplyr::rename(Z_AM_beta = exposure_ID_AM_IVW_beta) %>%
  dplyr::rename(Z_AM_se = exposure_ID_AM_IVW_se) %>%
  mutate(Z_prod = Z_sq_beta*Z_AM_beta) %>%
  mutate(Z_prod_var = variance_of_product(Z_sq_beta, Z_sq_se, Z_AM_beta, Z_AM_se)) %>%
  mutate(Z_prod_se = sqrt(Z_prod_var)) %>%
  mutate(same_trait_corr_minus_MR_est_se = sqrt(variance_of_sum(outcome_ID_IVW_se, outcome_ID_couple_r_se)))

@@

  theme_minimal() +
  ggplot2::labs(x = paste0("max C"),
                y = paste("Correlation - MR-estimate")) +
  theme(legend.title=element_blank()) +
  #geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  theme_ipsum()

@@

Lastly, we attempted to compute a MVMR sum of the effects of all potential confounders as shown in the DAG below and compared these C-statistics to the difference between phenotypic correlation and MR-estimates, however the estimates did not make sense so we decided not to pursue this analysis.

@@

```{r MVMR_C_plot, echo = FALSE, eval = FALSE, message=FALSE}

discrepentant_traits <- unique(corr_potential_trait_confounders_pos$outcome_ID)

@@

  corr_potential_trait_confounders_pos_i <- corr_potential_trait_confounders_pos %>% filter(outcome_ID == !!outcome_ID)

@@

  MVMR_result_org <- MVMR_result_org %>% mutate(beta_2_var = 4*beta^2*se^2+2*se^4) %>%
    mutate(beta_2_se = sqrt(beta_2_var)) %>% 
    left_join(corr_potential_trait_confounders_pos_i %>% dplyr::select(exposure_ID, exposure_ID_AM_IVW_beta, exposure_ID_AM_IVW_se), by = c("exposure" = "exposure_ID")) %>%
    dplyr::rename(MVMR_Z_beta = beta) %>%
    dplyr::rename(MVMR_Z_se = se) %>%
    dplyr::rename(MVMR_Z_pval = pval) %>%
    mutate(MVMR_Z_sq_beta = MVMR_Z_beta^2) %>%
    dplyr::rename(MVMR_Z_sq_se = beta_2_se) %>% 
    dplyr::rename(Z_AM_beta = exposure_ID_AM_IVW_beta) %>%
    dplyr::rename(Z_AM_se = exposure_ID_AM_IVW_se) %>%
    mutate(Z_prod = MVMR_Z_sq_beta*Z_AM_beta) %>%
    mutate(Z_prod_var = variance_of_product(MVMR_Z_sq_beta, MVMR_Z_sq_se, Z_AM_beta, Z_AM_se))

@@

# Single-trait causal effects in couples.

@@

AM_result_sig <-
  household_MR_summary_AM %>% filter(IVW_pval < 0.05 / num_tests_by_PCs)
AM_sig_traits <-
  household_MR_summary_AM %>% filter(IVW_pval < 0.05 / num_tests_by_PCs) %>% pull(exposure_ID) %>% unique()
AM_sex_spec_result_sig <-
  household_MR_binned_het_sex_spec %>% filter(same_trait == TRUE) %>% filter(exposure_ID %in% AM_sig_traits) %>% filter(sex_het_pval < 0.05 /
                                                                                                                          num_tests_by_PCs_AM_sig)

@@

- For each trait in the pipeline, we tested the causal effect within couples (i.e. the number of analyses corresponds to the number of traits (`r dim(traits_final)[1]`)). 
- The significant results after adjusting for multiple hypothesis testing are shown below. 
- `IVW_meta_beta` and `IVW_meta_pval` corresponds to the beta and p-value of the meta-analyzed MR across sexes, respectively (i.e. MR estimates were computed in each sex-seperately using sex-specific SNP-exposure and SNP-outcome results and then meta-analyzed).  
- After adjusting for multiple tests (p < `0.05/`r num_tests_by_PCs``), identified `r dim(AM_result_sig)[1]` significant assortative mating (same-trait) MR results (corresponding to the table below and table "BF_sig_same_trait_MR" in the excel document).

```{r, AM_sig, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), echo = FALSE, include = TRUE, message = FALSE}

@@

AM_result_sig_nom.DT <- AM_sig_sex_het_nom %>% dplyr::select(trait_description, trait_ID, exposure_sex, IVW_beta, IVW_se, IVW_pval, sex_het_pval) %>% unique()

AM_sensitivity <- bind_rows(household_MR_SNPmeta_sensitivity) %>% filter(same_trait)
AM_joint.DT <- left_join(AM_result_sig, AM_sensitivity) %>% 
  dplyr::rename(trait_description = exposure_description) %>%
  dplyr::rename(trait_ID = exposure_ID) %>%
  dplyr::select(trait_description, trait_ID, wt_median_summary, wt_median_pval, wt_mode_summary, wt_mode_pval)

non_sig_MR_mode <- AM_joint.DT %>% filter(wt_mode_pval > 0.05)
non_sig_MR_median <- AM_joint.DT %>% filter(wt_median_pval > 0.05)

## Heterogeneity stats.

## Sensitivity follow-up.

Of those traits that had a significant causal-effect among couples by IVW, we tested the MR weighted-mode and MR weighted-median method for consistent findings.

- We found `r dim(non_sig_MR_mode)[1]` trait(s) which was/were not significant using the weighted-mode approach (p > 0.05). 
- We found `r dim(non_sig_MR_median)[1]` trait(s) which was/were not significant using the weighted-mode approach (p > 0.05).

Results can be seen in the table below.

```{r, AM_sensitivity, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), echo = FALSE, include = TRUE}

DT::datatable(AM_joint.DT, width = "100%") %>% formatSignif(columns=c('wt_median_pval', 'wt_mode_pval'), digits=3)

@@

- For each AM MR ($X_{p} \sim X_i$), analyses were run in each sex separately. 
- Of the `r dim(AM_sig_traits)[1]` significant results shown above, we tested to see if there was any difference in AM MR estimates between sexes. 
- A similar adjustment for multiple hypothesis testing was applied as above, resulting in `r num_tests_by_PCs_AM_sig` number of effective tests. 
- After adjusting for multiple hypothesis testing (p < `0.05/`r num_tests_by_PCs_AM_sig``), `r dim(AM_sex_spec_result_sig)[1]` traits showed significant differences amongst sexes.
- The table below shows the *nominally significant* results at `p < 0.05` (of which there are `r dim(AM_result_sig_nom.DT)[1]/2`) (a slight variation of this table corresponds to "nominally_sig_sex_diff" in the excel document).
- The $p_{binomial}$ can be calculated as `1 – pbinom(15, 64, 0.05)`, and the fold enrichment as `15/(64*0.05)`.
- Found that female estimates are on average larger than males (result of the paired t-test shown below).

@@

nominally_sig_sex_differences <- AM_result_sig_nom.DT %>% pivot_wider(names_from = exposure_sex, values_from = c(IVW_beta, IVW_se, IVW_pval)) 
# write.table(nominally_sig_sex_differences, "docs/manuscript_tables/nominally_sig_sex_differences.txt", row.names = F, sep=",", quote = F)

t.test(nominally_sig_sex_differences$IVW_beta_male, nominally_sig_sex_differences$IVW_beta_female, paired = T)

@@

- For each AM MR ($X_{p} \sim X_i$), analyses were run in the full sample as well as in 5 roughly equal sized bins according to time couple had been together (`time_together_even_bins`) estimated using [time at household variable](https://biobank.ctsu.ox.ac.uk/showcase/field.cgi?id=699), and median age (`age_even_bins`). 
- Of the `r length(AM_sig_traits)` significant results shown above, we tested to see if there was any significant difference in AM MR estimates amongst the two grouping variables. 
- Specifically, the outcome data set was split into 5 bins, and the SNP-outcome effect was estimated in each bin separately (and each sex seperately). 
- These outcome-SNP effects were then used to generate bin-specific MR estimates, using the same SNP-exposure effects from Neale. 
- To estimate the difference among bins, we used two approaches: 
    1. The Cochran's Q test to test for heterogeneity in meta-analyses *(of which we obtained no significant results, so not presented in paper)*.
    2. Tested the significance of the slope of linear model of median bin (either `age` or `time together`) versus the bin-specific MR estimate.

@@

We also tested the phenotypic correlation among the different bins. Below are the traits that show a significant trend either with Pearson correlation (p < `r 0.05/num_tests_by_PCs`) (the table below corresponds to "pheno_binned_corrs" in the excel document).

@@

@@

pheno_corr_pearson_sig <- binned_pheno_corr_summary %>% filter(bin_slope_pval_pearson < 0.05/num_tests_by_PCs) %>% arrange(group)

pheno_corr_pearson_sig <- replace_Neale_ID(pheno_corr_pearson_sig, traits_corr2_filled, "trait", "trait_description") %>%
  dplyr::rename(trait_ID = trait) %>%
  dplyr::select(trait_description, trait_ID, everything())

## create panel figure 
pheno_corr_pearson_sig$branches_to_pull <- NA

for(i in 1:dim(pheno_corr_pearson_sig)[1]){

  trait <- pheno_corr_pearson_sig[i, "trait_ID"][[1]]
  for(k in 1:length(exposure_info)){
  temp <- exposure_info[[k]]
  if(temp[which(temp$Value=="trait_ID"), "Info"] == trait) {break}
  }

  pheno_corr_pearson_sig$branches_to_pull[i] <- k
}

for(i in 1:dim(pheno_corr_pearson_sig)[1]){

  k <- pheno_corr_pearson_sig$branches_to_pull[i] 
  binned_pheno_corrs_i <- tar_read(binned_pheno_corrs, branches = k)[[1]]
  exposure_info_i <- tar_read(exposure_info, branches = k)[[1]]
  pheno_figs <- create_binned_pheno_figs (binned_pheno_corrs_i, exposure_info_i, grouping_var)

  name <- paste0("exposure_", i)
  group <- pheno_corr_pearson_sig[i, "group"][[1]]
  fig_num <- ifelse(group == "age_even_bins", 1, 2)
  figs[[name]] <- pheno_figs[[fig_num]] #tar_read(binned_pheno_figs, branches = i)[[1]][[fig_num]]

pheno_binned_figures <- plot_grid(plotlist = figs, nrow=2, ncol = 2, byrow = T, labels="AUTO")

## format table 
pheno_corr_pearson_sig.DT <- pheno_corr_pearson_sig %>% dplyr::select(-branches_to_pull) %>%
  mutate(bin_type = case_when(group == "age_even_bins" ~ "age", 
                              group == "time_together_even_bins" ~ "time_together")) %>%
  dplyr::select(-group) %>% dplyr::select(trait_description, trait_ID, bin_type, everything())

format_round_cols <- colnames(dplyr::select_if(pheno_corr_pearson_sig.DT, is.numeric))

DT::datatable(pheno_corr_pearson_sig.DT, width = "100%") %>% formatRound(columns=format_round_cols, digits = 3)

@@

```{r, pheno_binned_figs, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), echo = FALSE, out.width = "200%", dpi=600}

pheno_binned_figures

@@

We tested the impact of confounding on genetic PCs, N/E coordinates and the following traits (however only presenting a selection in paper):

@@

```{r corr_due_to_confounding_ratio, eval = FALSE, echo = FALSE}

@@

```{r corr_due_to_confounding_ratio_causal, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), echo = FALSE}

confounder_traits_sub <- confounder_traits[c(1,2,4,5)]

corr_impact_by_traits_causal_summary <- corr_impact_by_traits_causal %>% group_by(exposure_ID) %>%
   dplyr::summarise(mean_ratio = mean(corr_due_to_confounding_ratio), median_ratio = median(corr_due_to_confounding_ratio), sd_ratio = sd(corr_due_to_confounding_ratio), number_of_traits = n())

corr_impact_by_traits_causal_summary <- replace_Neale_ID(corr_impact_by_traits_causal_summary, traits_corr2_filled, "exposure_ID", "outcome_description")
